A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-07-31 _manuscript/_tex/manuscript.tex
update website and commit changes to _freeze and _manuscript

\abstract{When planning experimental research, determining an
appropriate sample size and using suitable statistical models are
crucial for robust and informative results. The recent replication
crisis underlines the need for more rigorous statistical methodology and
adequately powered designs. Generalized linear mixed models (GLMMs)
offer a flexible statistical framework to analyze experimental data with
complex (e.g., dependent and hierarchical) data structures. However,
available methods and software for a priori sample size planning for
GLMMs are often limited to specific designs. Tailored data simulation
approaches offer a more flexible alternative. Based on a practical case
study, the current tutorial equips researchers with a step-by-step guide
and corresponding code for conducting tailored a priori sample size
planning with GLMMs. We not only focus on power analysis but also
explain how to use the precision of parameter estimates to determine
appropriate sample sizes. We conclude with an outlook on the increasing
importance of simulation-based sample size planning.}

@@

\par{ This is version 3 of our preprint published at
\url{https://doi.org/10.31234/osf.io/rpjem}. All materials (reproducible
manuscript, including R code and all simulation results) are available
in the project's repository on the Open Science Framework (OSF) at
\url{https://osf.io/dhwf4/}. A Quarto Manuscripts website is hosted at
\url{https://Timo-Ko.github.io/glmm_simulation_tutorial/}.  The authors
declare that there were no conflicts of interest concerning the
authorship or the publication of this article. This research was funded
by a grant from the Volkswagen Foundation (Grant No.~98525).  Florian
Pargent and Timo Koch contributed equally to this
manuscript. Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows: Florian
Pargent:   formal
analysis, methodology, visualization, writing, editing; Timo K.
Koch:   conceptualization, formal
analysis, writing, editing; Anne-Kathrin Kleine:   formal
analysis, editing; Eva Lermer:   conceptualization, funding
acquisition, editing; Susanne
Gaube:   conceptualization, supervision, editing}
\par{Correspondence concerning this article should be addressed to Timo
K. Koch, Institute of Behavioral Science \& Technology, University of
St.~Gallen, Torstrasse
25, St.~Gallen 9000, Switzerland, Email: timo.koch@unisg.ch}
}

@@

When planning experimental research, it is essential to determine an
appropriate sample size and use appropriate statistical models to
analyze the data to ensure that the results are robust and informative
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}). The
recent replication crisis in Psychology and other disciplines has
illustrated many challenges surrounding the reproducibility and
reliability of study findings
(\citeproc{ref-yarkoniGeneralizabilityCrisis2022}{Yarkoni, 2022}). As a
result, there is a growing need for more rigorous statistical
methodology and the adoption of adequately powered experimental designs.
Multiple easy-to-use software solutions exist for simple statistical
models and experimental designs (\citeproc{ref-R-pwr}{Champely, 2020};
\citeproc{ref-lakensSimulationBasedPowerAnalysis2021}{Lakens \&
Caldwell, 2021}). A priori sample size planning for more complex
research designs such as flexible generalized linear mixed models (GLMM)
is not covered by standard software solutions. Researchers willing to
use this framework will need to use data simulation. In the present
work, we provide a tutorial on how to determine adequate sample sizes by
performing tailored simulation-based sample size planning for GLMMs.
After introducing some theoretical background on sample size planning,
we review existing software solutions in R and discuss under which
circumstances tailored data simulations are necessary. We proceed by
describing the relevant steps and decisions involved in tailored data
simulation, illustrated in a case study.

@@

\multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{mixedpower}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.57in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Simulation-based}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{power}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{analysis;}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{based}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{on}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{‘}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{simr}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{’}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{package}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{no}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{yes}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{none}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.79in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Kumle}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{et}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{al.}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{(2021)}}}} \\

@@

When the estimand has been defined, the next step in the research
process is to write code that simulates the data-generating process of
the planned study. This requires specifying a generative process for all
predictor variables used in the final data analysis. Realistic
assumptions can be quite challenging for observational studies or
continuous predictor variables, which is beyond the scope of this
tutorial. However, this step is much easier for experimental studies
with only categorical predictor variables because the distribution of
predictors is fixed by the study design. When all predictor variables
have been simulated, one can use the structure of a suitable GLMM to
simulate the dependent variable. To simulate the GLMM, one requires
plausible values for all model parameters. We will discuss strategies on
how these values can be obtained later. Because we have full control
over the data-generating process in a tailored simulation, it is
possible to model specific aspects of the planned study, like data
missing by design or assuming that some subjects might drop out. The
quality of the results of the sample size planning crucially depends on
the plausibility of the simulated data-generating process. However, we
argue that even a strongly simplified data-generating process (e.g.,
only a small number of interaction effects; only random intercepts, and
no random slopes; assuming that data is missing completely at random)
can yield informative results.

@@

Population parameters are all model parameters estimated in a GLMM, in
particular the regression coefficients of the fixed effects and the
standard deviation of the random effects (and the correlation between
random effects in more complicated models). In the absence of previous
studies with the same design or pilot data, strategies to specify
population parameters require access to domain knowledge from domain
experts. Because most domain knowledge can only be expressed in
unstandardized measurement units of a specific application, we argue
that unstandardized effect sizes are usually preferable over
standardized effect sizes for tailored simulation-based sample size
planning. The basic idea of all strategies is to quantify or visualize
the data-generating process implied by certain values of population
parameters in an intuitive way that enables calibration of population
parameters based on the available knowledge of domain experts. Although
we use frequentist model estimation in our tutorial, many strategies
demonstrated in this chapter are inspired by research on monitoring the
plausibility of model assumptions in applied Bayesian statistics
(\citeproc{ref-gelmanBayesianWorkflow2020}{Gelman et al., 2020}).

@@

\subsubsection{Iterate with domain
experts}\label{iterate-with-domain-experts}

@@

Gathering domain knowledge and the specification of population parameter
values is an iterative process, in which domain experts repeatedly
assess the plausibility of simulated data. As a first step, domain
experts can be interviewed to ``elicit'' their domain knowledge about
how the data of the planned study is expected to look. As most domain
experts are no experts in statistical modeling and GLMMs, they often
struggle without further guidance to communicate their knowledge in a
way that is useful when specifying the parameters for data simulation.
For this reason, we suggest that after an initial interview, the analyst
who is familiar with the structure of the GLMM selects an initial set of
insightful descriptive statistics and model-based quantities. Then they
reenter into an iterative discussion with the domain experts where some
set of population values are selected, and the plausibility of resulting
implied quantities is evaluated. Then, the population parameters are
updated until the domain experts are satisfied with the result. During
this process, the monitored descriptive statistics and model-based
quantities can be updated or extended to capture as much available
domain knowledge as possible.

@@

Note that this does not always mean that one will specify the same GLMM
that was used when specifying the data-generating process. On the one
hand, using a simpler model for data simulation than for model
estimation can be a useful strategy in scenarios where making plausible
assumptions for complicated random effect structures and interactions is
not feasible. On the other hand, using a more complex model for data
simulation than for model estimation can be a useful strategy in
scenarios where one has specific domain knowledge about aspects of the
data-generating process that are still difficult to estimate with the
current state-of-the-art in multilevel modeling.

@@

When target power has been specified, the number of subjects and the
number of items in our study design can be traded against each other
based on practical considerations. Although it is recommended to justify
the power (\citeproc{ref-lakensSampleSizeJustification2022}{Lakens,
2022a}), we adopt the common heuristic to target a power of 0.8 to
detect an effect of the expected size implied by our data simulation.
This could be achieved by collecting data from 250 subjects (about 25\%
of which will be experts), each completing the same 50 items (with
advice present in about 67\% of cases, which is correct in about 80\% of
cases with present advice). If collecting data from 250 subjects is not
feasible, an alternative would be to recruit 200 subjects but increase
the length of the experiment to 70 items. However, 70 items might take
too long to complete for the radiologists participating in the study,
who have a busy schedule. The simulation suggests that it might also be
possible to plan a shorter experiment with only 30 items if it is
feasible to recruit an even higher number of subjects (to be determined
by additional rounds of power analysis). Design parameters that also
affect power, and which could be investigated in the simulation to find
a more optimal trade-off, are the ratio of experts to students, the
frequency of whether advice is presented at all, and whether it is
correct or incorrect.\footnote{An advanced alternative to our heuristic
  trade-off between subjects and items would be to explicitly optimize
  for cost-efficient study designs using the \emph{mlpwr} package
  (\citeproc{ref-zimmerSampleSizePlanning2023}{Zimmer et al., 2023}).
  The package performs simulation based power analysis in a surrogate
  modeling framework that can evaluate power for a large grid of design
  combinations more efficiently. The framework also requires users to
  write their own \texttt{sim\_and\_analyse} function and can be
  considered a great extension on what we cover in our tutorial.}

@@

In our case study, we have performed simulation-based sample size
planning from a single set of parameter values that reflect our
assumptions of an expected effect size. Instead of extracting this
expected effect size from meta-analyses or pilot data, which has been
the main focus of previous tutorials (e.g.,
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021}),
we have demonstrated some strategies to determine plausible parameter
values in GLMMs based on domain knowledge. When sample sizes are chosen
based on the results of our simulation-based power analysis, a future
study will be informative to reject the null hypothesis if an effect of
our \emph{expected size} is present (or estimate the effect with
satisfying precision). However, if the true effect is indeed smaller,
power (or precision) will be lower, and the study might not be
sufficiently informative. A common, more conservative strategy for
sample size justification is to perform sample size planning for the
smallest effect size of interest (SESOI). An effect smaller than the
SESOI would be considered too small to be interesting or practically
meaningful, even if the effect is not actually zero
(\citeproc{ref-kingPointMinimalImportant2011}{King, 2011};
\citeproc{ref-lakensEquivalenceTestingPsychological2018}{Lakens, Scheel,
et al., 2018}). Unfortunately, specifying a plausible SESOI is a
challenging task (for strategies on how to specify a SESOI, see Lakens
(\citeproc{ref-lakensSampleSizeJustification2022}{2022a}), Riesthuis
(\citeproc{ref-riesthuisSimulationBasedPowerAnalyses2024}{2024}), or
Lakens (\citeproc{ref-lakensImprovingYourStatistical2022}{2022b})). When
domain knowledge or formal theories about the research topic of interest
are too vague to specify a meaningful SESOI, it is still recommended to
demonstrate power or precision for different effect sizes in a
\emph{sensitivity power analysis}
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}). By
simulating power (or precision) for different effect sizes (in addition
to the different number of subjects and items), one can make sure that
power (or precision) would still be sufficient to detect smaller effect
sizes than our expected effect or at least get an impression of how
strongly power (or precision) depends on the size of the true effect.
For our case study that investigates combined hypotheses in a GLMM
modeling framework, sensitivity analysis would require manually
specifying additional sets of plausible parameter values that reflect
scenarios with smaller or larger differences between groups with respect
to our specific research question. Power (or precision) could then be
simulated for several of these scenarios (across different numbers of
subjects and items, as considered earlier). Although recommended, we do
not present a sensitivity analysis for our case study to keep the length
of the tutorial manageable.

@@

@@

@@

@@

@@

>>>>

\abstract{When planning experimental research, determining an
appropriate sample size and using suitable statistical models are
crucial for robust and informative results. The recent replication
crisis underlines the need for more rigorous statistical methodology and
adequately powered designs. Generalized linear mixed models (GLMMs)
offer a flexible statistical framework to analyze experimental data with
complex (e.g., dependent and hierarchical) data structures. However,
available methods and software for a priori sample size planning for
GLMMs are often limited to specific designs. Tailored data simulation
approaches offer a more flexible alternative. Based on a practical case
study where we focus on a binomial GLMM with two random intercepts and
discrete predictor variables, the current tutorial equips researchers
with a step-by-step guide and corresponding code for conducting tailored
a priori sample size planning with GLMMs. We not only focus on power
analysis but also explain how to use the precision of parameter
estimates to determine appropriate sample sizes. We conclude with an
outlook on the increasing importance of simulation-based sample size
planning.}

@@

\par{ This is version 3.5 of our preprint published at
\url{https://doi.org/10.31234/osf.io/rpjem}. All materials (reproducible
manuscript, including R code and all simulation results) are available
in the project's repository on the Open Science Framework (OSF) at
\url{https://osf.io/dhwf4/}. A Quarto Manuscripts website is hosted at
\url{https://Timo-Ko.github.io/glmm_simulation_tutorial/}.  The authors
declare that there were no conflicts of interest concerning the
authorship or the publication of this article. This research was funded
by a grant from the Volkswagen Foundation (Grant No.~98525).  Florian
Pargent and Timo Koch contributed equally to this
manuscript. Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows: Florian
Pargent:   formal
analysis, methodology, visualization, writing, editing; Timo K.
Koch:   conceptualization, formal
analysis, writing, editing; Anne-Kathrin Kleine:   formal
analysis, editing; Eva Lermer:   conceptualization, funding
acquisition, editing; Susanne
Gaube:   conceptualization, supervision, editing}
\par{Correspondence concerning this article should be addressed to Timo
K. Koch, Institute of Behavioral Science \& Technology, University of
St.~Gallen, Torstrasse
25, St.~Gallen 9000, Switzerland, Email: timo.koch@unisg.ch}
}

@@

When planning experimental research, it is essential to determine an
appropriate sample size and use appropriate statistical models to
analyze the data to ensure that the results are robust and informative
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}). The
recent replication crisis in Psychology and other disciplines has
illustrated many challenges surrounding the reproducibility and
reliability of study findings
(\citeproc{ref-yarkoniGeneralizabilityCrisis2022}{Yarkoni, 2022}). As a
result, there is a growing need for more rigorous statistical
methodology and the adoption of adequately powered experimental designs.
Multiple easy-to-use software solutions exist for simple statistical
models and experimental designs (\citeproc{ref-R-pwr}{Champely, 2020};
\citeproc{ref-lakensSimulationBasedPowerAnalysis2021}{Lakens \&
Caldwell, 2021}). A priori sample size planning for more complex
research designs such as flexible generalized linear mixed models (GLMM)
is not covered by standard software solutions. Researchers willing to
use this framework will need to use data simulation. In the present
work, we provide a tutorial on how to determine adequate sample sizes by
performing tailored simulation-based sample size planning for GLMMs.
After introducing some theoretical background on sample size planning,
we review existing software solutions in R and discuss under which
circumstances tailored data simulations are necessary. We proceed by
describing the relevant steps and decisions involved in tailored data
simulation, illustrated in a case study where we focus on a binomial
GLMM with two random intercepts and discrete predictor variables.

@@

\multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{mixedpower}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.57in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Simulation-based}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{power}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{analysis;}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{based}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{on}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{‘}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{lme4}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{’}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{package}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{no}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{yes}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{none}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.79in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Kumle}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{et}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{al.}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{(2021)}}}} \\

@@

When the estimand has been defined, the next step in the research
process is to write code that simulates the data-generating process of
the planned study. This requires specifying a generative process for all
predictor variables used in the final data analysis. Realistic
assumptions can be quite challenging for observational studies or
continuous predictor variables, which is beyond the scope of this
tutorial. However, this step is much easier for experimental studies
with only categorical predictor variables because the distribution of
predictors is fixed by the study design. When all predictor variables
have been simulated, one can use the structure of a suitable GLMM to
simulate the dependent variable. To simulate the GLMM, one requires
plausible values for all model parameters. We will discuss strategies on
how these values can be obtained later. Because we have full control
over the data-generating process in a tailored simulation, it is
possible to model specific aspects of the planned study, like data
missing by design or assuming that some subjects might drop out. The
quality of the results of the sample size planning crucially depends on
the plausibility of the simulated data-generating process. If the
simulated data generating process is less complex than the true process,
one can expect that sample size planning underestimates the required
sample size and the planned analysis has inflated type I error rates
(\citeproc{ref-matuschekBalancingTypeError2017}{Matuschek et al.,
2017}). However, Matuschek et al.
(\citeproc{ref-matuschekBalancingTypeError2017}{2017}) have also shown
in simulations with LMMs that while fitting \emph{maximal models}
(i.e.~that include all possible random effects) safeguards against
inflated type I error rates, this can lead to a great loss in power if
the variance in the random effects is actually small. Thus, we argue
that even a simplified data-generating process (e.g., only a small
number of interaction effects; only random intercepts, and no random
slopes; assuming that data is missing completely at random) that is only
plausible under idealized circumstances, can yield informative results
and is preferred over performing no systematic sample size planning.

@@

Population parameters are all model parameters estimated in a GLMM, in
particular the regression coefficients of the fixed effects and the
standard deviation of the random effects (and the correlation between
random effects in more complicated models). In the absence of previous
studies with the same design or pilot data, strategies to specify
population parameters will always require access to some domain
knowledge. In our experience, research teams for projects where tailored
simulation-based sample size planning is necessary typically consist of
\emph{analysts} (who are primarily responsible for study design and
statistical analysis), and \emph{domain experts} (who often provide the
original research question and are responsible for data collection). In
smaller projects, analysts and domain experts might be the same persons.
In these research settings, a broad review of the available literature
typically plays a more important role. Because most domain knowledge can
only be expressed in unstandardized measurement units of a specific
application, we argue that unstandardized effect sizes are usually
preferable over standardized effect sizes for tailored simulation-based
sample size planning. The basic idea of all strategies is to quantify or
visualize the data-generating process implied by certain values of
population parameters in an intuitive way that enables calibration of
population parameters based on the available domain knowledge. Although
we use frequentist model estimation in our tutorial, many strategies
demonstrated in this chapter are inspired by research on monitoring the
plausibility of model assumptions
(\citeproc{ref-gelmanBayesianWorkflow2020}{Gelman et al., 2020};
\citeproc{ref-schadPrincipledBayesianWorkflow2021}{Schad et al., 2021})
and eliciting prior information
(\citeproc{ref-bocktingSimulationbasedPriorKnowledge2024}{Bockting et
al., 2024}; \citeproc{ref-hartmannFlexiblePriorElicitation2020}{Hartmann
et al., 2020};
\citeproc{ref-mikkolaPriorKnowledgeElicitation2023}{Mikkola et al.,
2023}; \citeproc{ref-stefanPracticalChallengesMethodological2022}{Stefan
et al., 2022}) in applied Bayesian statistics.

@@

\subsubsection{Iterate with domain
expertise}\label{iterate-with-domain-expertise}

@@

Tailored simulation-based sample size planning always requires access to
domain knowledge. However, gathering domain knowledge and the
specification of population parameter values is not always
straightforward and can be better described as an iterative process, in
which the plausibility of simulated data is repeatedly compared with the
available domain expertise. As a first step, dedicated domain experts
(if available) can be interviewed to \emph{elicit} their domain
knowledge about how the data of the planned study is expected to look.
Because most domain experts are no experts in statistical modeling and
GLMMs, they often struggle without further guidance to communicate their
knowledge in a way that is useful when specifying the parameters for
data simulation. For this reason, we suggest that after an initial
interview, the analyst who is familiar with the structure of the GLMM
selects an initial set of insightful descriptive statistics and
model-based quantities. Then, they reenter into an iterative discussion
with the domain experts where some set of population values are
selected, and the plausibility of resulting implied quantities is
evaluated. The population parameters are updated until the domain
experts are satisfied with the result. During this process, the
monitored descriptive statistics and model-based quantities can be
updated or extended to capture as much available domain knowledge as
possible. Even if no dedicated domain experts are available, the basic
principles stay the same except that initial parameter values are often
derived from a more extensive literature review. Descriptive statistics
and model-based quantities are still used to evaluate whether the
implied assumptions about the data generating process are plausible.
Often, this judgment evolves over several iterations as the analyst
becomes more familiar with the model and the study design.\footnote{Developing
  more effective strategies on how to elicit and visualize domain
  knowledge is currently an active area of research, primarily in the
  context of Bayesian modeling
  (\citeproc{ref-bocktingSimulationbasedPriorKnowledge2024}{Bockting et
  al., 2024};
  \citeproc{ref-hartmannFlexiblePriorElicitation2020}{Hartmann et al.,
  2020}; \citeproc{ref-mikkolaPriorKnowledgeElicitation2023}{Mikkola et
  al., 2023};
  \citeproc{ref-stefanPracticalChallengesMethodological2022}{Stefan et
  al., 2022}). We expect that future advances in this field will also be
  highly relevant for tailored simulation-based sample size planning
  with frequentist models.}

@@

Note that this does not always mean that one will specify the same GLMM
that was used when specifying the data-generating process. On the one
hand, using a simpler model for data simulation than for model
estimation can be a useful strategy in scenarios where making plausible
assumptions for complicated random effect structures and interactions is
not feasible, or the omitted components are expected to have small
effects (\citeproc{ref-matuschekBalancingTypeError2017}{Matuschek et
al., 2017}). On the other hand, using a more complex model for data
simulation than for model estimation can be a useful strategy in
scenarios where one has specific domain knowledge about aspects of the
data-generating process that are still difficult to estimate with the
current state-of-the-art in multilevel modeling. In fact, applying
models that are more parsimonious than the true data generating process
can provide a better trade-off between type I error rates (or bias) and
power (or precision) in many scenarios
(\citeproc{ref-matuschekBalancingTypeError2017}{Matuschek et al.,
2017}).

@@

When target power has been specified, the number of subjects and the
number of items in our study design can be traded against each other
based on practical considerations. Although it is recommended to justify
the power (\citeproc{ref-lakensSampleSizeJustification2022}{Lakens,
2022a}), we adopt the common heuristic to target a power of 0.8 to
detect an effect of the expected size implied by our data simulation.
This could be achieved by collecting data from 250 subjects (about 25\%
of which will be experts), each completing the same 50 items (with
advice present in about 67\% of cases, which is correct in about 80\% of
cases with present advice). If collecting data from 250 subjects is not
feasible, an alternative would be to recruit 200 subjects but increase
the length of the experiment to 70 items. However, 70 items might take
too long to complete for the radiologists participating in the study,
who have a busy schedule. The simulation suggests that it might also be
possible to plan a shorter experiment with only 30 items if it is
feasible to recruit an even higher number of subjects (to be determined
by additional rounds of power analysis). Design parameters that also
affect power, and which could be investigated in the simulation to find
a more optimal trade-off, are the ratio of experts to students, the
frequency of whether advice is presented at all, and whether it is
correct or incorrect.\footnote{An advanced alternative to our heuristic
  trade-off between subjects and items would be to explicitly optimize
  for cost-efficient study designs using the \emph{mlpwr} package
  (\citeproc{ref-zimmerSampleSizePlanning2023}{Zimmer et al., 2023}).
  The package performs simulation-based power analysis in a surrogate
  modeling framework that can evaluate power for a large grid of design
  combinations more efficiently. The framework also requires users to
  write their own \texttt{sim\_and\_analyse} function and can be
  considered a great extension on what we cover in our tutorial.}

@@

In our case study, we have performed simulation-based sample size
planning from a single set of parameter values that reflect our
assumptions of an expected effect size. Instead of extracting this
expected effect size from meta-analyses or pilot data, which has been
the main focus of previous tutorials (e.g.,
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021}),
we have demonstrated some strategies to determine plausible parameter
values in GLMMs based on domain knowledge. When sample sizes are chosen
based on the results of our simulation-based power analysis, a future
study will be informative to reject the null hypothesis if an effect of
our \emph{expected size} is present (or estimate the effect with
satisfying precision). However, if the true effect is indeed smaller,
power (or precision) will be lower, and the study might not be
sufficiently informative. A common, more conservative strategy for
sample size justification is to perform sample size planning for the
smallest effect size of interest (SESOI). An effect smaller than the
SESOI would be considered too small to be interesting or practically
meaningful, even if the effect is not actually zero
(\citeproc{ref-kingPointMinimalImportant2011}{King, 2011};
\citeproc{ref-lakensEquivalenceTestingPsychological2018}{Lakens, Scheel,
et al., 2018}). Unfortunately, specifying a plausible SESOI is a
challenging task (for strategies on how to specify a SESOI, see Lakens
(\citeproc{ref-lakensSampleSizeJustification2022}{2022a}), Riesthuis
(\citeproc{ref-riesthuisSimulationBasedPowerAnalyses2024}{2024}), or
Lakens (\citeproc{ref-lakensImprovingYourStatistical2022}{2022b})). When
domain knowledge or formal theories about the research topic of interest
are too vague to specify a meaningful SESOI, it is still recommended to
demonstrate power or precision for different effect sizes in a
\emph{sensitivity power analysis}
(\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}). By
simulating power (or precision) for different effect sizes (in addition
to the different number of subjects and items), one can make sure that
power (or precision) would still be sufficient to detect smaller effect
sizes than our expected effect or at least get an impression of how
strongly power (or precision) depends on the size of the true effect.
For our case study that investigates combined hypotheses in a GLMM
modeling framework, sensitivity analysis would require manually
specifying additional sets of plausible parameter values that reflect
scenarios with smaller or larger differences between groups with respect
to our specific research question. Power (or precision) could then be
simulated for several of these scenarios (across different numbers of
subjects and items, as considered earlier). In steps 2 and 4 of the
tutorial, we have briefly discussed scenarios where the applied
statistical model is more (or less) parsimonious than the
data-generating process. Sensitivity analysis can be used to assess the
consequences of a mismatch by investigating different combinations of
data-generating processes and statistical models. Such extended
simulations can be a great resource to make an informed decision about
the concrete model to estimate for the planned study where the true
data-generating process is unknown. Although recommended, we do not
present a sensitivity analysis for our case study to keep the length of
the tutorial manageable.

@@

\bibitem[\citeproctext]{ref-bocktingSimulationbasedPriorKnowledge2024}
Bockting, F., Radev, S. T., \& Bürkner, P.-C. (2024). Simulation-based
prior knowledge elicitation for parametric {Bayesian} models.
\emph{Scientific Reports}, \emph{14}(1), 17330.
\url{https://doi.org/10.1038/s41598-024-68090-7}

@@

\bibitem[\citeproctext]{ref-hartmannFlexiblePriorElicitation2020}
Hartmann, M., Agiashvili, G., Bürkner, P., \& Klami, A. (2020). Flexible
prior elicitation via the prior predictive distribution. In J. Peters \&
D. Sontag (Eds.), \emph{Proceedings of the 36th conference on
uncertainty in artificial intelligence ({UAI})} (Vol. 124, pp.
1129--1138). PMLR.

@@

\bibitem[\citeproctext]{ref-mikkolaPriorKnowledgeElicitation2023}
Mikkola, P., Martin, O. A., Chandramouli, S., Hartmann, M., Abril Pla,
O., Thomas, O., Pesonen, H., Corander, J., Vehtari, A., Kaski, S.,
Bürkner, P.-C., \& Klami, A. (2023). Prior {Knowledge Elicitation}: {The
Past}, {Present}, and {Future}. \emph{Bayesian Analysis}, \emph{-1}(-1).
\url{https://doi.org/10.1214/23-BA1381}

@@

\bibitem[\citeproctext]{ref-schadPrincipledBayesianWorkflow2021}
Schad, D. J., Betancourt, M., \& Vasishth, S. (2021). Toward a
principled {Bayesian} workflow in cognitive science. \emph{Psychological
Methods}, \emph{26}(1), 103--126.
\url{https://doi.org/10.1037/met0000275}

@@

\bibitem[\citeproctext]{ref-stefanPracticalChallengesMethodological2022}
Stefan, A. M., Evans, N. J., \& Wagenmakers, E.-J. (2022). Practical
challenges and methodological flexibility in prior elicitation.
\emph{Psychological Methods}, \emph{27}(2), 177--197.
\url{https://doi.org/10.1037/met0000354}
